From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-21 Pi+/paper/finite.tex
Revising Lehmer section

\begin{definition}[$\downf : (n : \Nat) \to {\Fin[n]} \to {\List[\Fin[\suc[k + n]]]}$]
  \begin{align*}
    n \downf 0       & \defeq \nil                       \\
    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
  \end{align*}
\end{definition}

@@

To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will need to define functions back and forth between
the two types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect
to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
representatives. We know that these are the unique normal forms in the set-quotient given by $q \comp \normf$, but now
we will describe what these representatives look like exactly, using Lehmer codes.

@@

From grade school combinatorics, we know that there are $\fac{n}$ permutations
on a finite set with $n$ elements. The factorial function is defined by
recursion on natural numbers. However, now, for every $n$, we want to produce a
type, which is a finite set, with cardinality $\fac{n}$. And, to characterise
$\Aut[\Fin[n]]$, we further need to construct a bijection between this type and
$\Aut[\Fin[n]]$.

First, let's define this type with $\fac{n}$ elements, we name this type family
$\Lehmer : \Nat \to \UU$, which is defined by recursion on $\Nat$ as follows.
This is the obvious definition of factorials by recursion, but categorified from
natural numbers to sets.

\begin{definition}
  \begin{align*}
    \Lehmer[0]       & \defeq \unit                           \\
    \Lehmer[\suc[n]] & \defeq \Fin[\suc[n]] \times \Lehmer[n]
  \end{align*}
\end{definition}

\todo{Subexcedant sequences and factorial definitions are equivalent, explain
  this!}

The name Lehmer comes from Lehmer
codes~\cite{lehmerTeachingCombinatorialTricks1960} which are known in
Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960}. There are many
ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer
codes are a particularly convenient way to represent permutations on a
computer,~\review{they are compact and have exactly the right cardinality.
  $\Lehmer[n]$ is a $n+1$-element tuple, where the position $k \leq n$ has an
  element of $\Fin[k]$. The 0-th position is trivial, so we ignore it, and in
  both the example below and the Agda proof, consider only the remaining
  $n$-element tuple.}

\vc{This is just the classical algorithm to explain the example, not the actual
  type-theoretic proof.}

Suppose we have a permutation $p$ on an $n$-element set
$\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it as follows.
$\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of
inversions of the element $\el{k}$ in $p$, i.e. the number of elements smaller
than $\el{k}$ occurring after $\el{k}$.

As an example, consider the following tabulated presentation of the permutation:

@@

\[
  p =
  \begin{array}{ccccccccccccccc}
    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
    \hline                                                             \\
    | & \el{2} & | & \el{0} & | & \el{1} & | & \el{4} & | & \el{3} & | \\
    \hline                                                             \\
  \end{array}
\]

@@

Thus, the Lehmer code for the permutation $p$ is the 5-tuple
$l = (0, 0, 2, 0, 1)$.

@@

\begin{proposition}
  For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.
\end{proposition}

\begin{proposition}
  For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.
\end{proposition}

\begin{proposition}
  For any $n : \Nat$,
  \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}{k}} \).
\end{proposition}

@@

\begin{proof}
  For $n = 0$, note that $\Lehmer[0]$ is contractible, and so is $\Aut[\Fin[0]]$. For $n = \suc[m]$, we have the
  following chain of equivalences.
  \[\arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Aut[\Fin[\suc[m]]]                                                               \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \Fin[m]}                \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\Fin[m] \eqv \Fin[m]}                                    \\
      \eqv & \Fin[\suc[m]] \times \Aut[\Fin[m]]                                                \\
      \eqv & \Fin[\suc[m]] \times \Lehmer[m]                                                   \\
    \end{array}
  \]
\end{proof}

>>>>

\begin{definition}[$\downf : (n : \Nat) \to {\Fin[n]} \to {\List[\Fin[\suc[k + n]]]}$]
  \begin{align*}
    n \downf \zero   & \defeq \nil                       \\
    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
  \end{align*}
\end{definition}

@@

To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will need to define functions back and forth between
the two types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect
to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
representatives. We know that these are the unique normal forms in the set-quotient given by $\quotinc \comp \normf$,
but now we will describe what these representatives exactly look like, using an encoding called Lehmer
codes~\cite{lehmerTeachingCombinatorialTricks1960}.

@@

From grade school combinatorics, we know that there are $\fac{n}$ permutations on a finite set with $n$ elements. For
every $n$, we thus want to produce a finite set with cardinality $\fac{n}$, and a bijection with $\Aut[\Fin[n]]$. There
are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called "subexcedant sequences",
or "factoriadics", which give the factorial number system. They are a particularly convenient way of representing
permutations on a computer - since they have exactly the right cardinality, they are bitwise-optimal.

Formally, $\Lehmer[n]$ is an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
the 0-th position is trivial, in practice, it is ignored, and the sequence starts at
1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.\todo{There are more examples - listed
  in~\cite{duboisTestsProofsCustom2018}}. We can define $\Lehmer$ in two equivalent ways, by a simple recursion on
$\Nat$, or as a type family generated by two constructors.

\begin{definition}[$\Lehmer : \Nat \to \UU$]
  \begin{gather*}
    \begin{aligned}
      \Lehmer[\zero]   & \defeq \Fin[\suc[\zero]]                     \\
      \Lehmer[\suc[n]] & \defeq \Fin[\suc[\suc[n]]] \times \Lehmer[n]
    \end{aligned}
    \qquad
    \begin{aligned}
      \lzero & : \Lehmer[\zero]                                                     \\
      \lsuc  & : \forall n, r, (r \leq \suc[n]) \to \Lehmer[n] \to \Lehmer[\suc[n]]
    \end{aligned}
  \end{gather*}
\end{definition}

\todo{Explain why the two definitions are equivalent?}
\note{immersion here?}

Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
as follows. $\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of inversions of the element
$\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
following tabulated presentation of the permutation:

@@

\[
  \sigma =
  \begin{array}{ccccccccccccccc}
    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
    \hline                                                             \\
    | & \el{2} & | & \el{0} & | & \el{1} & | & \el{4} & | & \el{3} & | \\
    \hline                                                             \\
  \end{array}
\]

@@

Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 0, 2, 0, 1)$.

@@

We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$. \vc{These can be understood
  using simple counting arguments using the pigeonhole principle, but we have to construct the equivalences.}

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.
    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.
    \item For any $n : \Nat$,
          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}{k}} \).
  \end{enumerate}
\end{proposition}

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\zero]]$. For $n = \suc[m]$, we
  compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Lehmer[\zero]    \\
      \eqv & \unit             \\
      \eqv & \Aut[\Fin[\zero]] \\
    \end{array}
    \qquad\qquad
    \begin{array}{rl}
           & \Aut[\Fin[\suc[m]]]                                                               \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \Fin[m]}                \\
      \eqv & \dsum{k : \Fin[\suc[m]]}{\Fin[m] \eqv \Fin[m]}                                    \\
      \eqv & \Fin[\suc[m]] \times \Aut[\Fin[m]]                                                \\
      \eqv & \Fin[\suc[m]] \times \Lehmer[m]                                                   \\
      \eqv & \Lehmer[\suc[m]]
    \end{array}
  \end{gather*}
\end{proof}
